Measuring Interaction Design \\before Building the System: a Model-Based Approach

### 2016-04-11 paper.tex
implemented changes promised in the rebuttal

\numberofauthors{2}
\author{%
  \alignauthor{1st Author Name\\
    \affaddr{Affiliation}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}\\
  \alignauthor{2nd Author Name\\
    \affaddr{Affiliation}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}\\
}

@@

We present an approach that allows a designer to assess interaction
design qualities such as efficiency, error-proneness and
recovery from errors. Key importance is given to the ability to
\begin{inparaenum}[(a)]
\item understand how supportive a user interface (UI) is with respect
  to user efficiency;
\item understand how prone the UI is to user navigation
  errors;
\item understand how recoverable the UI is from those errors; and,
\item perform objective quantitative comparison of different designs
  to support construction and engineering of interactive systems.
\end{inparaenum}
All this can be done before developing prototypes of the UI.

To explain our work we present a comparison of four web mail front
ends. We have chosen this domain because it is easy to understand and
yet several questions cannot be easily answered. We applied the same
techniques also in other domains, such as HVAC and other embedded UIs.

Developing good UIs for web or mobile applications is a
complex and expensive endeavor. One reason is the combination of
devices, interaction modalities and workflows that need to be
supported.
%
Adopting Usage Centered Development (UCD) practices is effective, as
is following established design principles~\cite{constantine99}. In
particular, one of the most effective technique is early prototyping
to explore part of the five-dimensional fidelity prototyping
space~\cite{mccurdy06,buxton07}. This is particularly effective when
paired with usability investigations based on user testing or
heuristic evaluations~\cite{preece02,rubin08}.

@@

Our contribution consists of
\begin{inparaenum}[(a)]
\item the idea of using a UML state machine model to specify
  interaction scenarios,
\item the development of a tool (\migtool, Measuring Interaction
  Graphs Tool) that transforms models and scenario specifications into
  interaction paths, and
\item the definition of metrics that provide concise, precise and
  objective measures of a design.
\end{inparaenum}
The examples presented below show that among four web mail
applications, and with respect to a typical usage scenario, Gmail is
the most efficient and flexible UI, with the best ability to recover
from user errors, but only for users that possess a certain level of
proficiency. In fact, Gmail has the largest number of shortest paths
(when users are supposed to make 0 or 1 error); when users make 2 or
more errors the number of paths drops significantly, which indicates
that the error-proneness of the UI has reduced; in the best case,
Gmail features also the shortest paths, requiring 13\% fewer steps;
however, the probability that a user hits an optimal path with Gmail
is 10 times smaller than the best of the other applications, and the
probability that a random walker hits a state that is not due to an
error is 10\% smaller than the best of the other applications.
%
These values suggest that Gmail offers more ways to accomplish tasks
included in the scenario, that comparably more of these ways do not
involve extra steps, that they require fewer steps, and that it might
be more difficult for novice users to exploit the most efficient
ways. Further inspections show that some differences are due to the
slightly different interaction structures adopted for uploading
attachments. Thus, redesigning the methods with which competing
applications deal with attachments could improve their usability. If
Gmail didn't exist yet, by using \migtool\ its designers could obtain
these answers well before developing prototypes and performing
usability studies.

@@

In \cite{fayollas2014}, hierarchical task models (specified with
HAMSTERS) are used to specify how pilots are expected to interact with
airplane cockpit interfaces in tasks such as changing value to some
flight parameter. The paper in particular discusses how these task
models can be used to specify alternative ways to achieve te same
goal, and analyse the consequences on safety and reliability of the resulting
system. Compared to MIGTool, this approach cannot take advantage of a
model of the system and  instead relies only on models of the
tasks. The conseguence is that analysing alternative ways to achieve
the same goal requires developing different models; in addition,
exploring suitability of a given system with respect to certain tasks
is complex; and finally, hierarchical task models cannot take
advantage of contexts and rich control flows that are available with
state-transition representations, limiting therefore the richness of
the scenarios being analysed.

@@

In this section we provide a description of how \migtool\ computes
execution traces.  They are potential paths (\ie, sequences of
connected states of the model) that users can follow when performing a
given scenario.  The generation process encompasses the following
steps:
\begin{inparaenum}[(1)]
\item automatic flattening of the statechart model;
\item manual definition of the interaction scenario;
\item automatic generation of execution traces; and
\item interactive analysis of results.
\end{inparaenum}

@@

\migtool\ takes as input UML statecharts, which are a generalization
of finite state automata (FSA), with an expressive language that
includes hierarchical levels of abstraction, concurrent regions, states
and pseudostates, guards, and an extended state notion based on an
arbitrary underlying computational model.  Harel, the inventor of
statecharts, gives an interesting retrospective view of how they were
invented, back in early eighties, and why they were appealing also to
non experts~\cite{harel09}.%   For the sake of brevity, we refrain from
% describing them here in detail; the interested reader is referred to
% the UML standard~\cite{uml25} or some of the textbooks that deal with
% them, like~\cite{samek09}.

By using statecharts, the behavior of UIs can be represented by
associating states to screens and particular configurations of
widgets, and transitions to actions performed by users or by the
system itself. In the classification reported in~\cite{hilbert2000},
actions belong to the ``abstract interaction events'' category.

Because statechart models take advantage of abstraction features and a
rich set of connecting pseudostates, they are not suitable to be
directly processed to compute metrics. For this reason, \migtool\
first \emph{flattens} the model. Flattening is a process often used
when statecharts have to be automatically
processed~\cite{briand2005:test:uml}, and it means to produce a FSA
that is behaviorally equivalent to the original statechart, with no
hierarchy between states and no concurrent regions. In most cases this
leads to an exponential number of states and transitions in the FSA,
but because the process is completely automatic and there is no need
to manually inspect the resulting FSA, this aspect is in many cases
not relevant.  \migtool\ produces an XML representation (graphML) of
the resulting FSA, the \emph{interaction graph}. It is a directed
multigraph\footnote{A directed multigraph is a directed graph such
  that there are 2 or more edges that have the same end points. Cycles
  are paths that include 2 or more occurrences of the same
  vertex. Loops are edges that start and end on the same
  vertex. ``Geodesic path'' is a synonymous term with ``shortest
  path''.}, potentially with cycles and loops, with edges that are
labeled with the name of the corresponding action. In the following we
will be using as synonymous the terms state/vertex, and action/edge.

@@

Because in all but the most trivial interaction graphs there are loops
or cycles, the set of possible interaction traces is infinite. For
this reason, scenarios need to be defined and used as constraints on
the possible execution traces that can be generated.  Users of
\migtool\ define interaction scenarios by specifying key steps (called
\emph{bridge sets}) that users are expected to go through; we call
this process \emph{grounding usage scenarios on models}. Each bridge
set is specified by selecting a subset of the edges of the interaction
graph. In general a specification of a scenario includes an initial state and a
sequence of bridge sets. For example, to specify a scenario for
replying to an email message, one could select all the edges
associated to the action \texttt{reply} (bridge set 1), followed by
edges labeled with \texttt{typeBody} (bridge set 2), followed by edges
labeled with \texttt{send} (bridge set 3). A well formed bridge set
consists of 1 or more edges (an empty bridge set would make the
scenario unviable). In this way scenarios with cycles can be
formulated (\eg, reply twice to two messages). A \emph{stage} of a
scenario comprises two consecutive bridge sets.
%
To cope with multi edges, the interaction graph is \emph{simplified}:
all edges between a   pair of vertices are merged into a single one,
whose label includes the original ones.

\newpage
\subsection{Searching traces}

@@

Figure~\ref{fig:model} shows part of the model of Gmail. At some
point, a user is viewing all the conversations of the inbox (state
\texttt{viewingConversations}); available actions include moving to
the next or previous block of conversations, refreshing the list, or
opening a specific conversation (transition
\texttt{open(conversation)}). This transition (which is assumed to
occur when the user clicks on one of the visible conversations) leads
to the state \texttt{viewingAConversation}, where the behavior of the
system is defined by a more detailed state machine. By default the
user is viewing a conversation, but by performing the \texttt{reply}
action the UI moves to a state called \texttt{replying}, where the
body of the reply can be typed, the subject can be changed, or another
addressee can be added.

@@

Inspection of the interaction graph is not particularly useful because
even for small graphs like the one obtained from our Gmail model no
particular structure is evident that was not known from the
model.
%  Figure~\ref{fig:gom} shows a plot using a circular layout of
% the 47 states: 
Because of the large number of cycles that exist among states, edges
in the interaction graph form an intricate web of possible action
sequences. We believe this greatly reduces usefulness of
typical network analysis metrics, such as \emph{betweenness,
  eccentricity, page-rank, eigenvalue} centrality measures.

@@

To be able to obtain results that bear upon usability, we process
further the graph, by specifying scenarios and computing traces.  The
following fragment of R code specifies the bridge sets of the chosen
scenario: it comprises, in the given order, two occurrences of
\texttt{open(conversation)}, followed by \texttt{reply}, followed by
\texttt{typeBody}, etc.: \lstset{language=R}
\begin{lstlisting}
edgesWithLabel(gmail,"open(conversation)"),
edgesWithLabel(gmail,"open(conversation)"), 
edgesWithLabel(gmail,"reply"),
edgesWithLabel(gmail,"typeBody"),
edgesWithLabel(gmail,"send"),
edgesWithLabel(gmail,"compose"),
edgesWithLabel(gmail,"addRecipient"),
edgesWithLabel(gmail,"open(files)"),
edgesWithLabel(gmail,"writeBody"),
edgesWithLabel(gmail,"writeSubject"),
edgesWithLabel(gmail,"send")
\end{lstlisting}
In plain language such a scenario means opening a conversation,
opening a second one, replying to the last message of the
second conversation by typing the body of the response, sending it, and then
composing a new message by adding a recipient, an attachment, typing
the body and subject, and finally sending it.

@@

@@

UML state diagrams provide a very expressive language, well suited to
specify behavior of UIs based on discrete events. Evidence of this can
also be found in recent OMG standards, such as
IFML~\cite{ifml13}. Even though there are fundamental limits
(inability to handle undo/redo's, because that goes beyond a finite
state representation; inability to handle customizable toolbars,
because that requires models that change at runtime; inability to
handle perceptive UIs, because they are not well suited to be
described in terms of discrete states), in many practical cases they
can be be isolated and/or ignored~\cite{gb15:icwe}.

@@

A designer using \migtool\ is free from other concerns that in the end
affect usability, and the conclusions that are derived with \migtool\
can be combined with other results \emph{after} the analysis is
performed. As mentioned above, ours is an agile modeling approach
\cite{ambler02} that:
\begin{inparaenum}[(a)]
\item enables analysis of a UI well before it is developed;
\item  requires \emph{only} a model of its behavior;
\item  supports exploration of different tasks/scenarios;
\item enables grounding of tasks/scenarios through a
  platform-independent specification. As such, grounding can be viewed
  as the specification of a \emph{reification} relation between a
  \emph{task-and-domain-concepts} level to an
  \emph{abstract-user-interface} level (we are using terms of the
  CAMELEON reference model \cite{calvary:iwc03:camaleon}).
\end{inparaenum}
It is possible to extend \migtool\ so that also task models
(specified, for example, through CTT, UsiXML, HAMSTERS
\cite{mori02,limbourg04:usixml,martinie15}) can be processed and used
as specification of the scenarios to be analyzed. At that point,
\migtool\ could be used as a tool to explore suitability of tasks with
respect to a behavioral model of a user interface. This problem has
been addressed in \cite{martinie15} through a mechanism that relates
task models to source code annotations of the implementation of the
user interface. In this way it is possible to simulate execution of
the tasks in parallel with execution of the system and ensure
compatibility of the system with respect to tasks.

@@

% \migtool\ is part of a suite of model-based tools for designing,
% analyzing and testing user interfaces developed by the company
% ANONYMIZED. 
\migtool\ is implemented partly in Java (model processing)
and partly in R/iGraph. Development of a web-based UI of \migtool\ is
underway; it will be freely available for research purposes. Currently
\migtool\ reads UML models represented as XMI files; it could be
easily extended to handle also SCXML representations of statecharts \cite{scxml}.

@@

The method should not be used to draw final usability conclusions, as
it is devoid of any concerns dealing with what is presented to users,
how they could perceive that, understand that, and manipulate that.
But, because the approach can be applied  when just a
specification of the user interface is available, it supports
construction and engineering of interactive systems and could help in
iteratively improving a design before building any UI prototype.

%\bibliographystyle{SIGCHI-Reference-Format}
%\bibliography{gb,database}
%%% -*-BibTeX-*-
%%% Do NOT edit. File created by BibTeX with style
%%% ACM-Reference-Format-Journals [18-Jan-2012].

\begin{thebibliography}{00}

%%% ====================================================================
%%% NOTE TO THE USER: you can override these defaults by providing
%%% customized versions of any of these macros before the \bibliography
%%% command.  Each of them MUST provide its own final punctuation,
%%% except for \shownote{}, \showDOI{}, and \showURL{}.  The latter two
%%% do not use final punctuation, in order to avoid confusing it with
%%% the Web address.
%%%
%%% To suppress output of a particular field, define its macro to expand
%%% to an empty string, or better, \unskip, like this:
%%%
%%% \newcommand{\showDOI}[1]{\unskip}   % LaTeX syntax
%%%
%%% \def \showDOI #1{\unskip}           % plain TeX syntax
%%%
%%% ====================================================================

\ifx \showCODEN    \undefined \def \showCODEN     #1{\unskip}     \fi
\ifx \showDOI      \undefined \def \showDOI       #1{{\tt DOI:}\penalty0{#1}\ }
  \fi
\ifx \showISBNx    \undefined \def \showISBNx     #1{\unskip}     \fi
\ifx \showISBNxiii \undefined \def \showISBNxiii  #1{\unskip}     \fi
\ifx \showISSN     \undefined \def \showISSN      #1{\unskip}     \fi
\ifx \showLCCN     \undefined \def \showLCCN      #1{\unskip}     \fi
\ifx \shownote     \undefined \def \shownote      #1{#1}          \fi
\ifx \showarticletitle \undefined \def \showarticletitle #1{#1}   \fi
\ifx \showURL      \undefined \def \showURL       #1{#1}          \fi

\bibitem{ambler02}
{S. Ambler}. 2002.
\newblock {\em Agile Modeling: Effective Practices for eXtreme Programming and
  the Unified Process}.
\newblock Wiley.
\newblock
\showISBNx{0471202827}

\bibitem{appert2008swingstates}
{C. Appert} {and} {M. Beaudouin-Lafon}. 2008.
\newblock \showarticletitle{SwingStates: adding state machines to Java and the
  Swing toolkit}.
\newblock {\em Software: Practice and Experience\/} {38}, 11 (2008),
  1149--1182.
\newblock

\bibitem{conversy07}
{E. Barboni}, {S. Conversy}, {D. Navarre}, {and} {P. Palanque}. 2007.
\newblock \showarticletitle{Model-based engineering of widgets, user
  applications and servers compliant with ARINC 661 specification}.
\newblock In {\em Interactive Systems. Design, Specification, and
  Verification}. Springer, 25--38.
\newblock

\bibitem{scxml}
{J. Barnett} {and} {et al}. 2015.
\newblock {\em State Chart {XML (SCXML)}: State Machine Notation for Control
  Abstraction}.
\newblock {T}echnical {R}eport. W3C.
\newblock
\newblock
\shownote{\url{http://www.w3.org/TR/2015/REC-scxml-20150901}.}

\bibitem{bellamy2011cogtool}
{R. Bellamy}, {B. John}, {and} {S. Kogan}. 2011.
\newblock \showarticletitle{Deploying CogTool: integrating quantitative
  usability assessment into real-world software development}. In {\em Software
  Engineering (ICSE), 2011 33rd International Conference on}. IEEE, 691--700.
\newblock

\bibitem{gb15:icwe}
{G. Brajnik} {and} {S. Harper}. 2015.
\newblock \showarticletitle{Detaching control from data models in model-based
  generation of user interfaces}. In {\em Proc. of Int. Conf. on Web
  Engineering}, {F.~Frasincar} (Ed.). IEEE, Rotterdam, Netherlands.
\newblock

\bibitem{briand2005:test:uml}
{L.~C Briand}, {Y. Labiche}, {and} {J. Cui}. 2005.
\newblock \showarticletitle{Automated support for deriving test requirements
  from UML statecharts}.
\newblock {\em Software \& Systems Modeling\/} {4}, 4 (2005), 399--423.
\newblock

\bibitem{buxton07}
{B. Buxton}. 2007.
\newblock {\em User Experience: Getting the Design Right and the Right Design}.
\newblock Morgan Kaufmann.
\newblock

\bibitem{calvary:iwc03:camaleon}
{G. Calvary}, {J. Coutaz}, {D. Thevenin}, {Q. Limbourg}, {L. Bouillon}, {and}
  {J. Vanderdonckt}. 2003.
\newblock \showarticletitle{A Unifying Reference Framework for Multi-Target
  User Interfaces}.
\newblock {\em Interacting with Computers\/} {15}, 3 (2003), 289--308.
\newblock

\bibitem{ceri:00}
{S. Ceri}, {P. Fraternali}, {and} {A. Bongio}. 2000.
\newblock \showarticletitle{{Web Modeling Language (WebML)}: a modeling
  language for designing web sites}.
\newblock {\em Computer Networks\/}  {33} (2000), 137--157.
\newblock

\bibitem{constantine99}
{L.L. Constantine} {and} {L.A.D. Lockwood}. 1999.
\newblock {\em Software for use: a practical guide to the models and methods of
  usage-centered design}.
\newblock Addison-Wesley.
\newblock

\bibitem{fayollas2014}
{C. Fayollas}, {C. Martinie}, {P. Palanque}, {Y. Deleris}, {J-C. Fabre}, {and}
  {D. Navarre}. 2014.
\newblock \showarticletitle{An approach for assessing the impact of
  dependability on usability: application to interactive cockpits}. In {\em
  Tenth European Dependable Computing Conference (EDCC), 2014}. IEEE, 198--209.
\newblock

\bibitem{pirolli2007}
{W.-T. Fu} {and} {P. Pirolli}. 2007.
\newblock \showarticletitle{{SNIF-ACT}: A cognitive model of user navigation on
  the World Wide Web}.
\newblock {\em Human--Computer Interaction\/} {22}, 4 (2007), 355--412.
\newblock

\bibitem{gomez01}
{J. G\'{o}mez}, {C. Cachero}, {and} {O. Pastor}. 2001.
\newblock \showarticletitle{Conceptual Modeling of Device-Independent Web
  Applications}.
\newblock {\em IEEE MultiMedia\/} {8}, 2 (2001), 26--39.
\newblock
\showISSN{1070-986X}
\showDOI{%
\url{http://dx.doi.org/10.1109/93.917969}}

\bibitem{harel09}
{D. Harel}. 2009.
\newblock \showarticletitle{Statecharts in the making: a personal account}.
\newblock {\em CACM\/} {52}, 3 (March 2009), 67--75.
\newblock

\bibitem{heymann2007formal}
{M. Heymann} {and} {A. Degani}. 2007.
\newblock \showarticletitle{Formal analysis and automatic generation of user
  interfaces: approach, methodology, and an algorithm}.
\newblock {\em Human Factors: The Journal of the Human Factors and Ergonomics
  Society\/} {49}, 2 (2007), 311--330.
\newblock

\bibitem{hilbert2000}
{D.~M Hilbert} {and} {D.F. Redmiles}. 2000.
\newblock \showarticletitle{Extracting usability information from user
  interface events}.
\newblock {\em ACM Computing Surveys (CSUR)\/} {32}, 4 (2000), 384--421.
\newblock

\bibitem{hornbaek2006current}
{K. Hornb{\ae}k}. 2006.
\newblock \showarticletitle{Current practice in measuring usability: Challenges
  to usability studies and research}.
\newblock {\em International Journal of Human-Computer Studies\/} {64}, 2
  (2006), 79--102.
\newblock

\bibitem{horrocks99}
{I. Horrocks}. 1999.
\newblock {\em Constructing the User Interface with Statecharts}.
\newblock Addison-Wesley Longman Publishing Co., Inc., Boston, MA, USA.
\newblock
\showISBNx{0201342782}

\bibitem{john10}
{B.E. John}. 2010.
\newblock \showarticletitle{Reducing the Variability between Novice Modelers:
  Results of a Tool for Human Performance Modeling Produced through
  Human-Centered Design}. In {\em Proc. of the 19th Conf. on Behavior
  Representation in Modeling and Simulation}. Charleston, SC, USA, 95--102.
\newblock

\bibitem{koch02:uwe}
{N. Koch} {and} {A. Kraus}. 2002.
\newblock \showarticletitle{{The expressive power of {UML}-based web
  engineering}}. In {\em Second International Workshop on Web-Oriented Software
  Technology (IWWOST02)}, {D.~Schwabe}, {O.~Pastor}, {G.~Rossi}, {and}
  {L.~Olsina} (Eds.). 105--199.
\newblock

\bibitem{limbourg04:usixml}
{Q. Limbourg}, {J. Vanderdonckt}, {B. Michotte}, {L. Bouillon}, {and} {V.
  López~Jaquero}. 2004.
\newblock \showarticletitle{UsiXML: a Language Supporting Multi-Path
  Development of User Interfaces}. In {\em Proc. of 9th IFIP Working Conference
  on Engineering for Human-Computer Interaction jointly with 11th Int. Workshop
  on Design, Specification, and Verification of Interactive Systems,
  EHCI-DSVIS'2004} {\em (Lecture Notes in Computer Science)}, Vol. 3425.
  Springer-Verlag, Hamburg, Germany, 200--220.
\newblock

>>>>

\numberofauthors{2}
\author{%
  \alignauthor{Giorgio Brajnik\textsuperscript{1,2}\\
    \affaddr{\textsuperscript{1}Dip. di Matematica, Informatica e Fisica}\\
    \affaddr{Universit\`a di Udine, Italy}\\
    \email{brajnik@uniud.it}}\\
  \alignauthor{Simon Harper\textsuperscript{2}\\
    \affaddr{\textsuperscript{2}Computer Science School}\\
    \affaddr{University of Manchester, UK}\\
    \email{simon.harper@manchester.ac.uk}}\\
}

@@

We present an approach that allows a designer to assess interaction
design qualities such as efficiency, error-proneness and
recovery from errors. Key importance is given to the ability to
\begin{inparaenum}[(a)]
\item understand how supportive a user interface (UI) is with respect
  to user efficiency;
\item understand how prone the UI is to user navigation
  errors;
\item understand how recoverable the UI is from those errors; and,
\item perform objective quantitative comparison of different designs
  to support construction and engineering of interactive systems.
\end{inparaenum}
All this can be done before developing prototypes of the UI, similarly
to what is claimed and done in~\cite{silva14}: ``... the analysis of solutions in
the early phases of development is crucial as their results can then
be iteratively injected in the development cycle''.

To explain our work we present a comparison of four web mail front
ends. We have chosen this domain because it is easy to understand and
yet several questions cannot be easily answered. We applied the same
techniques also in other domains, such as HVAC (Heat, Ventilation and
Air Conditioning) and other embedded UIs.

Developing good UIs for web or mobile applications is a
complex and expensive endeavor. One reason is the combination of
devices, interaction modalities and workflows that need to be
supported.
%
Adopting Usage Centered Development practices is effective, as
is following established design principles~\cite{constantine99}. In
particular, one of the most effective technique is early prototyping
to explore part of the five-dimensional fidelity prototyping
space~\cite{mccurdy06,buxton07}. This is particularly effective when
paired with usability investigations based on user testing or
heuristic evaluations~\cite{preece02,rubin08}.

@@

Our contribution consists of
\begin{inparaenum}[(a)]
\item the idea of using a UML state machine model to support
  specification of interaction scenarios,
\item the development of a tool (\migtool, Measuring Interaction
  Graphs Tool) that, based on models, transforms scenario
  specifications into interaction paths, and
\item the definition of metrics that provide concise, precise and
  objective measures of the interaction structure entailed by a user
  interface. 
\end{inparaenum}
The examples presented below show that among four web mail
applications, and with respect to a typical usage scenario, Gmail is
the most efficient and flexible UI, with the best ability to recover
from user errors, but only for users that possess a certain level of
proficiency. In fact, Gmail has the largest number of shortest paths
(when users are supposed to make 0 or 1 error); when users make 2 or
more errors the number of paths drops significantly, which indicates
that the error-proneness of the UI has reduced; in the best case,
Gmail features also the shortest paths, requiring 13\% fewer steps;
however, the probability that a user hits an optimal path with Gmail
is 10 times smaller than the best of the other applications, and the
probability that a random walker hits a state that is not due to an
error is 10\% smaller than the best of the other applications.
%
These values suggest that Gmail offers more ways to accomplish tasks
included in the scenario, that comparably more of these ways do not
involve extra steps, that they require fewer steps, and that it might
be more difficult for novice users to exploit the most efficient
ways. Further inspections show that some differences are due to the
slightly different interaction structures adopted for uploading
attachments. Thus, redesigning the methods with which competing
applications deal with attachments could improve their usability. If
Gmail didn't exist yet, by using \migtool\ its designers could obtain
these answers well before developing prototypes and performing
usability studies.

@@

In \cite{fayollas2014}, hierarchical task models (specified with
HAMSTERS) are used to specify how pilots are expected to interact with
airplane cockpit interfaces in tasks such as changing value to some
flight parameter. The paper in particular discusses how these task
models can be used to specify alternative ways to achieve the same
goal, and analyses the consequences on safety and reliability of the
resulting system. Compared to MIGTool, this approach cannot take
advantage of a model of the system and instead relies only on models
of the tasks. The conseguence is that analysing alternative ways to
achieve the same goal requires developing different models; in
addition, exploring suitability of a given system with respect to
certain tasks is complex because no model of the system is
available. Finally, using \emph{only} hierarchical tasks as a modeling
language limits the richness of the scenarios being analysed, because
it is difficult to take advantage of contexts and rich control flows
that are available with state-transition representations, especially
with UML state diagrams.

@@

Traces are potential paths (\ie, sequences of
connected states of the model) that users can follow when performing a
given scenario.  The generation process encompasses the following
steps:
\begin{inparaenum}[(1)]
\item processing the model and automatic flattening of the statechart model;
\item manual definition of the interaction scenario;
\item automatic generation of execution traces; and
\item interactive analysis of results.
\end{inparaenum}

@@

\migtool\ takes as input UML statecharts, which are a generalization
of finite state automata (FSA) specified with an expressive language that
includes hierarchical levels of abstraction, concurrent regions, states
and pseudostates, guards, and an extended state notion based on an
arbitrary underlying computational model.  
% Harel, the inventor of
% statecharts, gives an interesting retrospective view of how they were
% invented, back in early eighties, and why they were appealing also to
% non experts~\cite{harel09}.%   For the sake of brevity, we refrain from
% % describing them here in detail; the interested reader is referred to
% the UML standard~\cite{uml25} or some of the textbooks that deal with
% them, like~\cite{samek09}.

By using statecharts, the behavior of UIs can be represented by
associating states to screens and particular configurations of
widgets, and transitions to actions performed by users or by the
system itself~\cite{horrocks99}. In the classification reported
in~\cite{hilbert2000}, actions belong to the ``abstract interaction
events'' category.

Because statechart models take advantage of abstraction features and a
rich set of connecting pseudostates, they are not suitable to be
directly processed to compute metrics. For this reason, \migtool\
first \emph{flattens} the model. Flattening is a process often used
when statecharts have to be automatically
processed~\cite{briand2005:test:uml}, and it means to produce a FSA
that is behaviorally equivalent to the original statechart, with no
hierarchy between states and no concurrent regions. In most cases this
leads to an exponential number of states and transitions in the FSA,
but because the process is completely automatic and there is no need
to manually inspect the resulting FSA, this aspect is in many cases
not relevant. Because \migtool\ does not use executable UML,
transition guards are not given semantics and are simply treated as
part of the event of a transition. \migtool\ produces an XML
representation (graphML) of the resulting FSA, the \emph{interaction
  graph}. It is a directed multigraph\footnote{A directed multigraph
  is a directed graph such that there are 2 or more edges that have
  the same end points. Cycles are paths that include 2 or more
  occurrences of the same vertex. Loops are edges that start and end
  on the same vertex. ``Geodesic path'' is a synonymous term with
  ``shortest path''.}, potentially with cycles and loops, with edges
that are labeled with the name of the corresponding action. In the
following we will be using as synonymous the terms state/vertex, and
action/edge.

@@

Because in all but the most trivial interaction graphs there are loops
or cycles, the set of possible interaction traces is infinite. For
this reason, scenarios need to be defined and used as constraints on
the possible execution traces that can be generated.  Users of
\migtool\ define interaction scenarios by specifying key steps (called
bridge sets) that users are expected to go through; we call
this process \emph{grounding usage scenarios on models}. A \emph{bridge set}
is a subset of the edges of the interaction graph; a well formed
bridge set is a non-empty set (an empty bridge set would make the
scenario unviable). In general a specification of a scenario includes
an initial state and a non-empty sequence bridge sets. For example, to
specify a scenario for replying to an email message, assuming an
initial state that corresponds to a user interface that displays all
messaes in the inbox, one could select all the edges associated to the
action \texttt{reply} (bridge set 1), followed by edges labeled with
\texttt{typeBody} (bridge set 2), followed by edges labeled with
\texttt{send} (bridge set 3).  In this way scenarios with cycles can
be formulated (\eg, reply twice to two messages). A \emph{stage} of a
scenario comprises two consecutive bridge sets.
%
To cope with multi edges, the interaction graph is \emph{simplified}:
all edges between a   pair of vertices are merged into a single one,
whose label includes the original ones.

\subsection{Searching traces}

@@

Figure~\ref{fig:model} shows part of the model of Gmail. At some
point, a user is viewing all the conversations of the inbox (state
\texttt{viewingConversations}); available actions include moving to
the next or previous block of conversations, refreshing the list, or
opening a specific conversation (transition
\texttt{open(conversation)}). This transition is assumed to occur when
the user clicks on any one of the visible conversations - it is if
there is a collection of transitions, each one for a different
conversation that can be opened; the parameter \texttt{conversation}
is bound to any occurrence of a conversation. Such transitions lead to the state
\texttt{viewingAConversation}, where the behavior of the system is
defined by a more detailed state machine. By default the user is
viewing a conversation, but by performing the \texttt{reply} action
the UI moves to a state called \texttt{replying}, where the body of
the reply can be typed, the subject can be changed, or another
addressee can be added.

@@

Inspection of the interaction graph is not particularly useful because
even for small graphs like the one obtained from our Gmail model no
particular structure is evident that was not known from the
model.
%  Figure~\ref{fig:gom} shows a plot using a circular layout of
% the 47 states: 
Because of the large number of cycles that exist among states, edges
in the interaction graph form an intricate web of possible action
sequences. We believe this greatly reduces usefulness of
typical network analysis metrics, such as \emph{betweenness,
  eccentricity, page-rank} and \emph{eigenvalue} centrality measures.

@@

To be able to obtain results that bear upon usability, we process
further the graph, by specifying scenarios and computing traces.  The
scenario, which assumes \texttt{ViewingConversations} is the initial
state, is specified with the following 11 bridge sets:
\newpage
\lstset{language=R}
\begin{lstlisting}
1  open(conversation)
2  open(conversation)
3  reply
4  typeBody
5  send
6  compose
7  addRecipient
8  open(files)
9  writeBody
10 writeSubject
11 send
\end{lstlisting}
In plain language such a scenario means opening a conversation, doing
something and then opening a second one, replying to the last message
of the second conversation by typing the body of the response, sending
it, and then composing a new message by adding a recipient, an
attachment, typing the body and subject, and finally sending
it. Notice that because the model precludes the possibility that there
are two \emph{consecutive} actions called \texttt{open(conversation)},
MIGTool has to find traces such that between bridge set 1 and 2 there
is at least one \texttt{goBack} action.

@@

One of the order 0 geodesic paths is shown below; each line
represents an action that the user is expected to perform; actions are
only those entailed by the model. Actions marked with ``*'' are those not
specified in the bridge sets and found in the interaction graph. 
\lstset{language=Java}
\begin{lstlisting}
1  open(conversation) 
2  goBack *
3  open(conversation) 
4  reply 
5  typeBody 
6  send 
7  compose 
8  attachFiles *
9  addRecipient[first] 
10 open(files) 
11 writeBody 
12 writeSubject 
13 send 
\end{lstlisting}
% \begin{tabular}{rp{7cm}}
%     1 & \texttt{open(conversation)} \\
%     2 & \texttt{goBack} \\ 
%     3 & \texttt{open(conversation)} \\
%     4 & \texttt{reply} \\
%     5 & \texttt{reply, addAddressee, typeBody, changeSubject} \\
%     6 & \texttt{cancel, send, open(message)[msgIsClosed], close (message) [msgIsOpen]}\\
%    7 & \texttt{compose} \\
%     8 & \texttt{attachFiles} \\
%     9 & \texttt{addRecipient[first]} \\
%     10  & \texttt{open(files)} \\
%     11 & \texttt{writeSubject, writeBody} \\
%     12 & \texttt{writeSubject, close(message)[msgIsOpen], writeBody,
%          open(message)[msgIsClosed]} \\
%     13 & \texttt{cancel, send} \\
%   \end{tabular}
%   %

@@

UML state diagrams provide a very expressive language, well suited to
specify behavior of UIs based on discrete events. in \migtool\ all the
UML language~\cite{uml25} for state machine has been used, including sundry
pseudostates. Evidence of this can also be found in recent OMG
standards, such as IFML~\cite{ifml13}. Even though there are
fundamental limits (inability to handle undo/redo's, because that goes
beyond a finite state representation; inability to handle customizable
toolbars, because that requires models that change at runtime;
inability to handle perceptive UIs, because they are not well suited
to be described in terms of discrete states), in many practical cases
they can be be isolated and/or ignored~\cite{gb15:icwe}. One
particular limitation of \migtool\ stems from the fact that it does
not rely on an executable UML metamodel; as a consequence behaviors
that rely on the fact that firing a transition broadcasts new events
cannot be properly represented and analyzed.

@@

A designer using \migtool\ is free from other concerns that in the end
affect usability, and the conclusions that are derived with \migtool\
can be combined with other results \emph{after} the analysis is
performed. As mentioned above, ours is an agile modeling approach
\cite{ambler02} that:
\begin{inparaenum}[(a)]
\item enables analysis of a UI well before it is developed;
\item  requires \emph{only} a model of its behavior;
\item  supports exploration of different tasks/scenarios;
\item enables grounding of tasks/scenarios through a
  platform-independent specification. As such, grounding can be viewed
  as the specification of a \emph{reification} relation between a
  \emph{task-and-domain-concepts} level to an
  \emph{abstract-user-interface} level (we are using terms of the
  CAMELEON reference model \cite{calvary:iwc03:camaleon}).
\end{inparaenum}
It is possible to extend \migtool\ so that also task models
(specified, for example, through CTT, UsiXML, HAMSTERS
\cite{mori02,limbourg04:usixml,martinie15}) can be processed and used
as specification of the scenarios to be analyzed. For example, a CTT
model could be annotated by associating subtasks to bridge sets and
then used to generate a sequence of bridge sets, from which
\migtool\ could generate traces and metrics.  At that point, \migtool\
could be used as a tool to explore suitability of tasks with respect
to a behavioral model of a user interface. This problem has been
addressed in \cite{martinie15} through a mechanism that relates task
models to source code annotations of the implementation of the user
interface. In this way it is possible to simulate execution of the
tasks in parallel with execution of the system and ensure
compatibility of the system with respect to tasks.

@@

% \migtool\ is part of a suite of model-based tools for designing,
% analyzing and testing user interfaces developed by the company
% ANONYMIZED. 
\migtool\ is implemented partly in Java (model processing)
and partly in R/iGraph~\cite{r14}. 
%Development of a web-based UI of \migtool\ is
%underway; it will be freely available for research purposes. 
Currently
\migtool\ reads UML models represented as XMI files; it could be
easily extended to handle also SCXML representations of statecharts \cite{scxml}.

@@

The method should not be used to draw final usability conclusions, as
it is devoid of any concerns dealing with what is presented to users,
how they could perceive, understand, and manipulate that.
But, because the approach can be applied  when just a
specification of the user interface is available, it supports
construction and engineering of interactive systems and could help in
iteratively improving a design before building any UI prototype.
\newpage
\bibliographystyle{SIGCHI-Reference-Format}
\bibliography{gb,database}

\balance{}
